$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-29 paper/paper.tex
Rearrange things, give multiplicities and typing contexts another pass

\todo{rename Scoped to Process in code}
\todo{make sure all citations are there}
\todo{change colors, color functions}

@@

A \emph{resource-aware} typing system for the \emph{\picalc{}} is distinctive in several ways.

\begin{description}
  \item [As a resource-aware typing system,] the language is constrained not just by types, but also by their usage annotations.
    We must be able to add up and split usage contexts, to do so we introduce a generalized algebra in \S \ref{multiplicities}.
    We provide an independent inductive definition of typing contexts and usage contexts in \S \ref{contexts}.
    \todo{why independent}

  \item [As a typing system for the \picalc{},] the terms (processes) in the typing judgment do not come accompanied by a type -- processes have no type.
    It is the communication channels that are annotated with types and usages.
    Typing judgments assert that the process correctly uses the channels in the context under which it is typed.
\end{description}

In this paper, we combine typing and annotation judgments into a single judgment.
Although the typing rules are for now completely syntax-directed, adding support for branching would introduce choice in the typing derivations.
If we were to keep typing judgments and annotation judgments separate, we would need a coherence relation making sure that the same choice is taken by both.
We avoid doing so by combining them into a single judgment.

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 or 1.
In this paper we generalize over linearity and provide an algebra on multiplicities that is capable of encoding shared and gradual types as well.
The algebra has a single binary operator $(\cdot)$ that crushes multiplicities together.
It must satisfy the following constraints:
\begin{description}

\item [Partiality]
  Not all multiplicities are necessarily compatible: in the linear \picalc{} $1$ cannot be split into $1$ and $1$.
\item [Functionality]
  Given two multiplicities $y$ and $z$, it must be computable whether a third multiplicity $x$ exists such that $\op{x}{y}{z}$.
  If such an $x$ exists, it must be unique.

\item [Injectivity]
  We need to be able to \emph{subtract} multiplicities.
  To do so, we rely on $\op{x}{y}{z}$ uniquely determining $y$ for any $x$ and $z$.

\item [Associativity]
  As a direct consequence of the structural congruence rule $\comp{P}{\comp{Q}{R}} \equiv \comp{\comp{P}{Q}}{R}$.

\item [Commutativity]
  As a direct consequence of the structural congruence rule $\comp{P}{Q} \equiv \comp{Q}{P}$.

\item [Distinguished elements]
  There must be a neutral element $0$ absorbed on both sides.
  The neutral element $0$ must also be the minimal element.
  There must be a special element $1$ to record usage.
\end{description}

\begin{nidefinition}
  Together, the aforementioned constraints give raise to a \emph{partial commutative monoid}.
  We deal with partiality by modeling the monoidal operation as a terniary relation.
  An algebra on the carrier set $C$ forms a $\Quantifier_C$ iff it satisfies the fields in Figure \ref{fig:multiplicities}.

@@

\begin{note}
  We will often work with input and output multiplicities pairs, and thus we use the notation $\Carrier^2$ to stand for the type $\Carrier \times \Carrier$, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all of its properties are trivially lifted.
\end{note}

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

\subsubsection{Example Type Systems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Add typing derivations for example processes}

\item [Shared]

The carrier $C$ is implemented as a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
All other laws are trivially satisfied.

\item [Linear]

The carrier $C$ is implemented as a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$), and is uninhabited in every other case.
All other rules follow trivially.

\item [Gradual]

The carrier $C$ is implemented as the type of natural numbers $\N$.
The element $\zero$ corresponds 0, and $\one$ to 1.
The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$.
All other rules follow trivially from the algebraic rules for the addition of naturals.
\todo{OD: in addition to the above, could we have examples of typing a process in each of the above type systems? It can be the same process everywhere and we can show how the typing changes...}
\end{description}

@@

In a resource-aware \picalc{} type multiplicities change every time input and output occurs; however, the types themselves remain constant.
If multiplicities need not to depend on types, keeping types and multiplicities in independent contexts allows for a more modular approach to the algebra on contexts of multiplicities.
In this paper we choose that approach, which entails accepting that base non-channel types have dedicated input and output multiplicities too -- we will take care of impeding that any actual input or output can occur through them in the typing rules.

\begin{nidefinition}
  Figure \ref{fig:prectx} defines types as either unit types, base types or channel types.
  Unit types $\unit$ provide a proof of inhabitance for types, needed for the $\new \PO_{\suc n} \eqeq \PO_n$ rule in subject congruence.

  Base types $\base{n}$ use a natural number $n$ that can be reified into a type by the host language.
  Using natural numbers instead of type universes is convenient: there is no need to litter type types with universe polymorphism.

  Channel types $\channel{t}{x}$ use $t$ as the type of data sent over the channel and $x$ as the capability of the data sent over the channel.
  When data is received on a channel of type $\channel{t}{x}$, the type $t$ will be appended to the context of types $\PreCtx_n$ and the capability $x$ to the context of capabilities $\Ctx_{idxs}$.

  Types form length-indexed \emph{pre-contexts} $\PreCtx_n$.

@@

\begin{nidefinition}
  To allow different multiplicity algebras to coexist we first define a length-indexed context of carrier indices $\Idxs_n$ (see \ref{fig:ctx}).
  On top of it, we define a context of multiplicities $\Ctx_{idxs}$.
  We lift the algebra on multiplicities to multiplicity contexts by constraining the algebra on multiplicity contexts to act on  contexts defined on the same underlying context of carrier indices.
  We use two multiplicities over a common index to represent multiplicities for both the input and the output capabilities ($\Carrier_{idx}^2$).

@@

@@

    \inferrule
    {\opsquared{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

@@

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplities exhausted, such that $\types{\gamma , t}{\Gamma , m}{P}{\Psi , \ell_{\o}}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma , t}{\Gamma , m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi , m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in Theorem \ref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
          \types{\gamma , t}{&\Psi , m}{P &&}{\Delta , \ell_{\o}} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma , t}{& \Gamma , m}{P &&}{\Theta , \ell_{\o}} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\ell_o$ witness, and merge with $\ell_i$ witness into $\ell_\#$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi , m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma , t}{\Gamma , m}{lower_0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi , m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}

>>>>

\todo{rename Scoped to Process in code}
\todo{make sure all citations are there}
\todo{change colors, color functions}
\todo{do not distinguish fields from functions, it is not worth it}

@@

Typing systems for the \picalc{} ensure that processes make a correct use of their channels.
Typing judgments are of the form $\Gamma \vdash P$ (for some typing context $\Gamma$ and some process $P$) --- types only appear in the contexts of typing judgments.
A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in \S \ref{multiplicities}.
We define independent typing contexts and usage contexts in \S \ref{context}.
We provide a typing judgment for a resource-aware \picalc{} in \S \ref{leftover-typing}.

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
In this paper we generalize the type of multiplicities by defining an algebra any such type must comply with.
Shared, gradual and linear usage types all satisfy these algebraic laws.

\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in Figure \ref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

@@

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for the type $\Carrier \times \Carrier$, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
\end{note}

Shared, gradual and linear multiplicity types can all be defined as an instance of the properties in Figure \ref{fig:multiplicities}.
For examples of how these multiplicity types are used in typing derivations, refer to \S \ref{example-derivations}.

\begin{description}
  \item [Shared]
    The carrier $C$ is a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
    Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
    The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
    All other laws are trivially satisfied.
  \item [Gradual]
    The carrier $C$ is implemented as the type of natural numbers $\N$.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$.
    All other rules follow trivially from the algebraic rules for the addition of naturals.
  \item [Linear]
    The carrier $C$ is implemented as a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$), and is uninhabited in every other case.
    All other rules follow trivially.
\end{description}

@@

In a resource-aware \picalc{}, usage annotations are transformed as a result of input and output; types howeverr are preserved.
In our case, usage annotations do not depend on types, and keeping types and usages in separate contexts allows for a more modular approach in our proofs.

\subsubsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{nidefinition}
  Figure \ref{fig:prectx} defines types as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types (we can trivially obtain it), and is used to such end in subject congruence --- for the $\new \PO_{\suc n} \eqeq \PO_n$ rule.
  The base type $\base{n}$ uses a natural number $n$ to represent types.
  This natural number can then be interpreted as a type in the host language.
  Using natural numbers instead of type universes is convenient: less universe polymorphism poluting specifications.

  The channel type $\channel{t}{x}$ uses $t$ to stand for the type of data that is being transmitted over the channel, and $x$ to stand for what usages of that data are transmitted.

  We keep track of types in a length-indexed \emph{pre-context} $\PreCtx_n$.

@@

\subsubsection{Usage Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}
\begin{nidefinition}
  To allow different multiplicity algebras to coexist we first define a length-indexed context of carrier indices $\Idxs_n$ (see \ref{fig:ctx}).
  On top of it, we define a context of multiplicities $\Ctx_{idxs}$.
  We lift the algebra on multiplicities to multiplicity contexts by constraining the algebra on multiplicity contexts to act on  contexts defined on the same underlying context of carrier indices.
  We use two multiplicities over a common index to represent multiplicities for both the input and the output capabilities ($\Carrier_{idx}^2$).

@@

%%%% REARRANGE
In this paper, we combine typing and annotation judgments into a single judgment.
Although the typing rules are for now completely syntax-directed, adding support for branching would introduce choice in the typing derivations.
If we were to keep typing judgments and annotation judgments separate, we would need a coherence relation making sure that the same choice is taken by both.
We avoid doing so by combining them into a single judgment.
%%%%

@@

    \inferrule
    {\opsquared{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{\constr{zero}}{t}{y}{\Gamma \comma z}}

@@

\subsubsection{Example derivations}

\todo{OD: in addition to the above, could we have examples of typing a process in each of the above type systems? It can be the same process everywhere and we can show how the typing changes...}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplities exhausted, such that $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \ell_{\o}}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in Theorem \ref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
          \types{\gamma \comma t}{&\Psi \comma m}{P &&}{\Delta \comma \ell_{\o}} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma \comma t}{& \Gamma \comma m}{P &&}{\Theta \comma \ell_{\o}} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\ell_o$ witness, and merge with $\ell_i$ witness into $\ell_\#$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{lower_0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi \comma m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}
